Probabilistic Catalogs from Crowded Star Fields

### 2012-08-31 paper/ms.tex
Random writing

However, standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, uncertainty should be
taken into account -- instead of simply estimating the position and flux of each
object, we should fully describe the fact that sometimes we are not certain of
the position and flux of each source.

Essentially, the creation of a catalog is an attempt to answer the question
``given the image we have obtained, what objects are present in the field and
what are their properties?''. This motivates a probabilistic (Bayesian)
approach to making catalogs.

\citet{2011MNRAS.415.3462F} have attempted this, by replacing the complex
$N$-component model with a single-component model, and using the
single-component
results to reconstruct what the $N$-component results would have been. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap) and also does not allow for unknown $N$,
which is the general situation. What we really need is a variable dimension
model, where $N$ is an unknown quantity to be inferred from the data.
The computational implementation of these models will require tools such as reversible jump Markov Chain Monte Carlo \citep{rjmcmc}.

\section{Bayesian Inference}
To quantitatively model uncertainties, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters (denoted
collectively by $\theta$) and we expect to obtain some data $D$. Our prior
knowledge (or uncertainty) about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta)
\end{equation}
We also model how the parameters give rise to the data, via a {\it sampling
distribution}:
\begin{equation}
p(D|\theta)
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions, one for each possible value of $\theta$. Note that
the sampling distribution is also an assertion of a prior state of knowledge.
In this case it models our prior information about the fact that the parameters
$\theta$ are going to influence the data in some way (otherwise this data
would be irrelevant).

When specific data $D^*$ are known, the prior gets updated to the posterior
via Bayes' rule:
\begin{eqnarray}
p(\theta|D=D^*) &\propto& p(\theta)p(D|\theta)|_{D=D^*} \\
&=& p(\theta)\mathcal{L}(\theta; D)
\end{eqnarray}
The modulating function is the {\it likelihood function}, which is the prior
probability of obtaining the actual data set $D^*$ as a function of the
parameters. As suggested by the above notation, it is obtained from the
sampling distribution with the actual data substituted in.

>>>>

However, standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, uncertainty should be
taken into account. Instead of simply {\it estimating} the position and flux of
each object in the image, we should fully describe the fact that sometimes we
are not certain of the position and flux of each source.

Essentially, the creation of a catalog is an attempt to answer the question
``given the image we have obtained, what objects are present in the field and
what are their properties?''. This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single single answer to this question (i.e. a single catalog), but rather a
probability distribution over the space of possible catalogs. This is
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image is itself unknown. Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
catalogs.

Bayesian object detection has been implemented by \citet{2011MNRAS.415.3462F}
under the assumption of a known $N$. This assumption is required for the
{\tt MultiNest} sampler to be applicable. Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if a variable-$N$ simulation had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N$ is an unknown quantity to be inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}.

\section{Bayesian Inference}
To quantitatively model uncertainties, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta)
\end{equation}
We also model how the parameters give rise to the data, via a {\it sampling
distribution}:
\begin{equation}
p(x|\theta)
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions, one for each possible value of $\theta$. Note that
the sampling distribution is also an assertion of a prior state of knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

When specific data $x^*$ has been obtained, our state of knowledge about $\theta$
gets updated from the prior to the posterior,
via Bayes' rule:
\begin{eqnarray}
p(\theta|x=x^*) &\propto& p(\theta)p(x|\theta)|_{x=x^*} \\
&=& p(\theta)\mathcal{L}(\theta; x)
\end{eqnarray}
The modulating function is the {\it likelihood function}, which is the prior
probability of obtaining the actual data set $x^*$ as a function of the
parameters\footnote{In the case that the sampling distribution is a probability
density, the likelihood is the probability density at the observed data. This
is usually fine, although one should be aware of the Borel-Kolmogorov ``paradox''}.
As suggested by the above notation, it is obtained from the
sampling distribution with the actual data substituted in.
